```{r setup, echo = F}
library(knitr); library(lattice)
opts_knit$set(out.format = 'html')
opts_chunk$set(highlight = TRUE, tidy = FALSE, fig.align = 'center', fig.height = 5.5, dev = 'svg')
```

..article: fill

<h1>
  <span class = 'yellow'>How to Do it with R</font><br/>
  <span class = 'white' style = 'font-size: 0.7em;'>
    A Tutorial
  </span><br/>
  <span class = 'white' style = 'font-size: 0.5em;'>
    by Ramnath Vaidyanathan
  </span>
</h1>

![slides](http://goo.gl/EpXln)

---

### Overview ###

The objective of this tutorial is to give you a quick overview of how to analyze data with `R`. We will focus our attention on the following topics

1. Working with Data
2. Gra


---

### Data ###


We will work with the built-in data set `mtcars`.

```{r mtcars1, echo = F}
head(mtcars)
```

You can inspect the data set in many ways. 

```{r mtcars-inspect, eval = F}
View(mtcars)       # view data in rstudio
head(mtcars)       # view first six rows
head(mtcars, 10)   # view first ten rows
tail(mtcars, 20)   # view last twenty rows
NROW(mtcars)       # view number of rows 
NCOL(mtcars)       # view number of columns
dim(mtcars)        # view rows x columns
str(mtcars)        # view structure of data
```

---

# Histograms #

---

### Basic Histogram ###

``` {r simple}
histogram(~ mpg, data = mtcars)
```

---

### Control Number of Bins ###

``` {r nint}
histogram(~ mpg, data = mtcars, nint = 20)
```

---

### Change Color ###

``` {r col}
histogram(~ mpg, data = mtcars, col = 'darkred') # see colors() for more
```

---

### Add x-Axis and y-Axis Labels ###

``` {r labels}
histogram(~ mpg, data = mtcars, xlab = 'Miles per Gallon')
```

---

### Split by Single Categorical Variable ###

``` {r single-cat}
histogram(~ mpg | cyl, data = mtcars)
```

---

### Split by Two Categorical Variables ###

``` {r two-cat}
histogram(~ mpg | cyl + gear, data = mtcars)
```

---

### Histogram for a Subset of the Data ###

``` {r hist-sub}
histogram(~ mpg, data = mtcars, subset = (gear == 4))
```

---

### Histograms: Summary ###

Here is a short summary of the commands you have seen on histograms.

``` {r mtcars-hist, eval = F}
hist(~ mpg, data = mtcars)                       # histogram of mpg
hist(~ mpg, data = mtcars, nint = 20)            # 20 bins
hist(~ mpg, data = mtcars, col = 'darkred')      # colored darkred.
hist(~ mpg | cyl, data = mtcars)                 # split by cyl
hist(~ mpg | cyl + gear, data = mtcars)          # split by cyl and gear
hist(~ mpg, data = mtcars, subset = (gear == 4)) # specific subset
```

---

# Box and Whisker Plots #


---

### Simple Boxplot ###

```{r bwplot}
bwplot(~ mpg, data = mtcars)
```

---

### Split by Single Categorical Variable ###

```{r bwplot1}
bwplot(~ mpg | cyl, data = mtcars)
```

---

### Side by Side Boxplot ###

The boxplots drawn by `bwplot` are horizontal by default. Hence the `y` axis displays the categorical explanatory variable, while the `x` axis displays the numerical response variable

```{r bwplot2, fig.height = 4.5}
bwplot(cyl ~ mpg, data = mtcars)
```

---

### Boxplot for a Subset of the Data ###

```{r bwplot3, fig.height = 4.5}
bwplot(cyl ~ mpg, data = mtcars, subset = (gear == 4))
```

---


# Scatterplots #

---

### Basic Scatterplot ###

```{r scplot1}
xyplot(mpg ~ wt, data = mtcars)
```

---

### Solid Circles as Points ###

```{r scplot2}
xyplot(mpg ~ wt, data = mtcars, pch = 16) # change point shapes using pch
```

---

### Change Color of Points ###

```{r scplot3}
xyplot(mpg ~ wt, data = mtcars, col = 'maroon', pch = 16)
```

---

### Add xAxis and yAxis Labels ###


```{r scplot4}
xyplot(mpg ~ wt, data = mtcars, col = 'maroon', pch = 16, 
  xlab = 'Weight', ylab = 'Miles Per Gallon')
```

---

### Color Points by Categorical Variable ###


```{r scplot5}
xyplot(mpg ~ wt, data = mtcars, group = cyl, auto.key = TRUE, pch = 16, 
  xlab = 'Weight', ylab = 'Miles Per Gallon')
```

---

### Split by Single Categorical Variable ###

```{r scplot6}
xyplot(mpg ~ wt | cyl, data = mtcars, col = 'maroon', pch = 16, 
  xlab = 'Weight', ylab = 'Miles Per Gallon')
```

---

### Split by Multiple Categorical Variable ###

```{r scplot7}
xyplot(mpg ~ wt | cyl + gear, data = mtcars, col = 'maroon', pch = 16, 
  xlab = 'Weight', ylab = 'Miles Per Gallon')
```

---

### Scatterplot for a Subset of the Data ###

```{r scplot8}
xyplot(mpg ~ wt, data = mtcars, subset = (cyl == 6))
```

---

# Subset #

---

### Working with Subsets of Data ###

Very often you might want to work with a smaller subset of the data. `R` makes it very easy to select subsets by combining conditional operations. The main operations are

    == : equal to
    != : not equal to
    >= : greater or equal to
    <= : lesser or equal to
    &  : and
    |  : or
    
Below, you will find these operators applied to choose specific subsets of the `mtcars` data set.

```{r eval = F, subset}
subset(mtcars, gear == 4)           # four gears
subset(mtcars, cyl != 2)            # not 2 cylinders
subset(mtcars, mpg > 20)            # mpg more than 20
subset(mtcars, mpg > 20 & wt < 20)  # mpg > 20 AND wt < 20
subset(mtcars, mpg > 20 | wt < 20)  # mpg > 20 OR wt < 20
```







